---
title: "Playing with Enterotypes"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'enterotypes.html'))})
author: "Donny Hoang"
date: "4/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal: Work more independently within R.

Today's goal is the same as last week's: for you to become comfortable working in R on your own. We will begin with using sample data from phyloseq, and will work within phyloseq to make things easier.

On Wednesday, we discussed reanalyzing the enterotype dataset within phyloseq to see if we accept the authors' definition of "enterotype."

## Excercises

### Load packages, data, and familiarize yourself with the dataset.

```{r}
library(vegan)
library(ggplot2)
library(phyloseq)
library(RColorBrewer)

data("enterotype")
```

You should now have the "enterotypes" phyloseq object loaded into your environment.

1) How many samples do you have? You can use the nsamples() function.
2) How many OTUs do you have? You can use the ntaxa() function.
3) What are some of the variables in the sam_data file? You can use the sample_variables() function.

This data set has some "NA" values in the "Enterotype" variable that will complicate downstream calculations. It also contans many unassigned oTUs. Let's remove those data.

```{r}
ent<-subset_samples(enterotype, !is.na(Enterotype))
ent <- subset_taxa(ent, Genus != "-1")
```

### Create and ordinate distance matrices

In the ordination lesson, we created a Bray-Curtis dissimilarity matrix, and used NMDS to ordinate. I ran into trouble trying to ordinate with NMDS, and some brief googling suggested it was because the dataset contains too many zeros and/or many species were observed only once or too few times between samples. We will ordinate with MDS/PCoA instead, using a variety of dissimilarity matrices. The paper we read used Jensen-Shannon dissimilarity. Let's compare it to Bray-Curtis (which we've used and it considers abundance), and Jaccard (which only considers presence/absence).

Let's try Jensen-Shannon first. Ordinate the data with "PCoA" as your ordination method and "jsd" as your dissimilarity/distance. Once the ordination has finished, plot it and color the data in by "SeqTech" and shape it by "Enterotype".

4) From eyeballing the graph, which variable better explains the data?

Let's do some stats to make more definitive conclusions. You'll need to use the function adonis() from the vegan package for this. Additionally, you will need to write out the sam_data from within the phyloseq object into a separte data frame. Use the code below to do that. Remember that adonis only works on a distance matrix, so you will need to create a separate distance matrix. Run your analysis of variance on both the "SeqTech" and the "Enterotype" variable.

```{r}
metadata<-as(sample_data(ent), "data.frame")
```

5) Which variable better describes the data? What does this result suggest to you about the concept of enterotypes?

Repeat your analysis above with both Bray-Curtis and Jaccard dissimilarity matrices. You will specify "bray" and "jaccard", respectively, in your code. Creating the plots is optional, you can just run the variance analysis.

6) How do the results compare between the dissimilarity matrices? Are your conclusions the same? What does this result suggest to you about the concept of enterotypes?
7) Bray-Curtis dissimilarity is typically used when abundance information is available. Jaccard is used when only presence/absence data is availabile. How does your Bray-Curtis and Jaccard results compare? Conceptually, what's different when you only consider presence/absence vs. abundance?

Hopefully you've concluded that SeqTech better describes this dataset than Enterotype. If you haven't, let me know.

### Subset the data by SeqTech, and reevaluate

Let's look within a specific SeqTech dataset. Subset the data to your favorite SeqTech variable, using the command subset_samples()

Repeat your plotting and statistical analysis above, using "jsd" and "bray" for your dissimilarity matrices.

8) How well is the data explained by the "Enterotype" variable? What do you conclude about the enterotype concept now?


As Marge suggested on Wednesday, we should take a closer look the community composition as well. Since there are 500+ taxa in this dataset, subset down to the top 20 before plotting the abundances with a bar plot. Facet your plot by "Enterotype".

9) Does your abundance bar plot support your ordination plot?
10) The enterotypes paper claimed that enterotypes could be based on the abundance of Bacteroides, Prevotella, and Ruminoccocus. Which enterotype corresponds to which genus?
11) How do you currently feel about the enterotype concept? What data would you need to support or contradict your current conclusions?


