---
title: "Lesson 1: R basics"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'pH_lesson.html'))})
author: "Donny Hoang"
date: "3/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Goal:
The goal of this lesson is to familiarize yourself with basic R commands. We will do achieve this goal by graphing pH data from our CR communities.


## Dataset:
We recently completed an experiment where we expanded one community (8A transfer 44) into six communities: A, B, C, D, E, F. There is an uninoculated negative control, X. These six communities were used to inoculate CR manipulated into six different treatments: CR pH 6, CR pH 6 + MES Buffer, CR pH 6 + 5% glucose, CR pH 6 + 10% glucose, CR pH 8, and CR pH 10. Thus, each treatment has six replicates and the replicates are communities A, B, C, etc. These communities were sampled daily for seven days, and we measured the pH of (almost) every sample.

Don't forget to download “SGH_CR_pH.csv” which is our data file.

## Excercise:

# Part 1: load in packages and your data
Create a new R script by navigating to “File” -> “New File” -> “R Script” save and name it as you like, but make sure the extension is “.R”

Good practice includes adding some basic information to our script:

```{r}
#Donny Hoang
#3.19.2020
#Graphing pH data
```

Your name, data, and a small blurb about what the script is for is the bare minimum in my opinion (but if I'm completely honest I don't always follow this.) You may notice that I have a "#" before the text. This symbol tells R not to run whatever text is on that line and is useful for taking notes throughout your code.

We will need the packages: ggplot2 and ggpubr. You can load in a package by typing:

```{r}
## if the packages are not installed, use these commands to install them
#install.packages('ggpubr')
#install.packages('ggplot2')
library(ggplot2)
library(ggpubr)
```

ggplot2 is what we will use to graph the data. ggpubr requires another package, magrittr (which should automatically load). magrittr has an operator we will use later on, "%>%", and ggpubr is useful for for organizing multiple graphs.

Load in your data. It should look something like this, depending on where the file is located on your computer.

```{r}
data <- read.csv("SGH_CR_pH.csv", header=TRUE)
```

Here, we are reading in our data, and storing it into an aptly named object called "data". Of course, you can name it whatever you want. I'm usually lazy and stick with "d". We are loading in a .csv file with the command read.csv(). This file has a header, so you must write "head=TRUE". What happens to your data if you write "header=FALSE"? You can check by typing:

```{r}
head(data)
# Read more about head by typing ?head
str(data) ## has a little more information about the columns class
# What's the difference between a Factor, integer, and numeric class?
summary(data)
```

Don't forget to fix your code before moving forward!


# Part 2: Organizing and sorting the data

There are A LOT of comparisons that we could make in this dataset and graphing ALL the data onto the same graph can be overwhelming and obstruct interpretation. For now, let’s focus on specific groups to make our analysis easier. Let’s focus on one treatment at a time. We will only graph data belonging to the CR pH 6 treatment. How can we tell R to only graph a specific portion of our data?

R has a nice function, subset(), which we will use to pull out only data that matches the treatment of interest. For example, if I only want CR pH 6 data I will type:

```{r}
data1 <- subset(data, data$treatment=="pH 6")
head(data1)
summary(data1)
```

In the code above, we are storing a subset of "data" into a new object "data1". Here, we have specified R to only add in data if the "treatment" column matches the string "pH 6" which represents our CR pH 6 treatment.

# Part 3: Graphing the data
We have now subset our data from one treatment. Let’s graph it!


```{r}
p1 <-ggplot(data=data1, aes(x=day, y=pH, group=replicate)) + ## initialize our plotting object
  ylim(4, 10) + ## set the y-axis limits
  geom_point(aes(shape=replicate)) + ## specify that we what to plot points
  geom_line() + ## specify that we also want lines
  theme_classic() ## make it pretty
p1 ## return the plotting object
```

Whoops, looks like we have an error. Since we have more groups (our replicates in this case) than R’s default shape palette, we must explicitly specify more than 6 shapes. If you look at the legend, you may notice how our "X" replicate does not have a symbol to represent it. This error can be fixed easily. Google “geom_point shapes” to get an idea of what symbols or shapes ggplot2 has to offer. I think making shapes identical to the replicate (that is, replicate A will have shape “A”) makes the most sense.

```{r}
shapeorder <- c(65, 66, 67, 68, 69, 70,  4)
```

This code creates an object, that contains a list of the exact shapes we want to use. Now we just need to make a small adjustment to our graph’s code. I added "scale_shape_manual()" to the code, and specified the "shapeorder" object we wrote above.

```{r}
p1 <- ggplot(data=data1,aes(x=day,y=pH,group=replicate))+
  ylim(4,10) +
  geom_point(aes(shape=replicate))+
  scale_shape_manual(values=shapeorder)+ ## what shapes we want to use
  geom_line() +
  theme_classic()
p1
```

Review what each piece of the command does. What is being added to the graph after every “+”? If you aren’t sure, delete a portion and see how that changes your graph.

You probably noticed that most of these are aesthetic changes, which are up to personal preference. I almost always use “theme_classic()” to remove grid lines and the gray background. I also try to avoid using default ggplot2 colors and try to ensure my color palettes are colorblind friendly, but that’s not relevant in this case.

Now we have some data graphed. It’s time to interpret!

1)	How does the pH change over time in this pH 6 treatment?
2)	Are the changes consistent between replicates? Can you think of any reasons that would explain the variation?

# Part 4: Writing a function to graph remaining treatments.

We only have one treatment graphed. We could copy+paste our graph code five more times and make manual edits, but manual edits potentially introduce errors. Also, that would make the script look ugly. This step is quite easy, as you’ve done most of the writing already. We simply need to write a function that takes in whatever data we feed it (denoted by "..."), and pipes it (denoted by %>%) to the function we wrote earlier. That is, the "%>%" takes whatever is on the lefthand side, "..." in this case, and puts it through whatever is on the righthand side. In this case, it's our graphing code.

```{r}
myplotmaker<- function(...){
  return(... %>%
           ggplot(aes(x=day,y=pH,group=replicate))+
           ylim(4,10) +
           geom_point(aes(shape=replicate))+
           scale_shape_manual(values=shapeorder)+
           geom_line() +
           theme_classic())
}
```


Now, we have a function that will create a graph of the data we give it. Feed it the previously subsetted “data1” by typing:

```{r}
p1 <- myplotmaker(data1)
p1
```

This plot should be identical to the graph you created before. Now, we need to subset the remaining data, and plot it. I have also added a small code to title each individual graph. 

```{r}
p1    <- myplotmaker(data1) + ggtitle("CR pH 6")

data2 <- subset(data, data$treatment=="pH 6 + MES")
p2    <- myplotmaker(data2) + ggtitle("CR pH6 + MES Buffer")

data3 <- subset(data, data$treatment=="pH 6 + 5% glucose")
p3    <- myplotmaker(data3) + ggtitle("CR pH6 + 5% glucose")

data4 <- subset(data, data$treatment=="pH 6 + 10% glucose")
p4    <- myplotmaker(data4) + ggtitle("CR pH6 + 10% glucose")

data5 <- subset(data, data$treatment=="pH 8")
p5    <- myplotmaker(data5) + ggtitle("CR pH 8")

data6 <- subset(data, data$treatment=="pH 10")
p6    <- myplotmaker(data6) + ggtitle("CR pH 10")
```

Our graphs are generated. You can look at them individually, but I think looking at all of them together will be more informative. Let's use the ggarrange() function from the package ggpubr. Can you figure out what each piece of the code is doing?

```{r}
ggarrange(p1, p2, p3, p4, p5, p6,
          ncol=3, nrow=2,
          legend="none")
```

You probably got a few errors about missing values. That's because I wasn't able to measure the pH of every single sample. You can ignore those messages.

Let’s interpret the data.

1)	The uninoculated control in each data set is replicate “X”. How does it change over the week? Is our data still reliable? What does this data suggest about the CR media itself?
2)	What conclusions can draw from each treatment? How does the pH change in each one? Are the changes consistent between treatments?
3)	How does the variation between replicates compare between treatments? What do you think this means?


Congrats! You made it through the first lesson. Hopefully, you feel a bit more familiar with the R language. It's OK if you don't feel completely comfortable yet. We will work on that together.